Mixed-Effects Model for MTLD by Gender

### 2018-08-19 paper.Rmd
minor formatting changes to text/code/plots

```{r load_packages, echo = FALSE, warning=FALSE}
library(papaja)
library(tidyverse)
library(childesr)
library(directlabels)
library(png)
library(here)
library(lme4)
library(ggthemes)
library(optimx)
knitr::opts_chunk$set(cache = TRUE, echo = FALSE)
```

@@

png::readPNG(here::here("images/schema.png")) %>% grid::grid.raster()
```

@@

`childes-db` is stored as a MySQL database, an industry-standard, open-source relational database server that can be accessed directly from a wide range of programming languages.
The childes-db project provides hosted, read-only databases for direct access and for `childesr` (described below) as well as compressed .sql exports for local installation.
While the former is appropriate for most users, local installation can provide performance gains by allowing a user to access the database on their machine or on their local network, as well as allowing users to store derived information in the same database.

@@

```{r, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# plot error between clan counts and our counts for speaker unigrams
freq_data <- read_csv("data/freq_testing.csv")

ggplot(freq_data %>%
         mutate(error = abs(clan - db) / clan),
       aes(error)) +
  geom_density() +
  xlim(0,.25)
```

@@

ggplot(mlu_data %>%
         mutate(error = abs(clan - db) / clan),
       aes(error)) +
  geom_density() +
  xlim(0,.25)
```

@@

# Interfaces for Accessing childes-db

We first discuss the `childes-db` web apps and then introduce the `childesr` `R` package.

@@

The ability to easily browse and explore the CHILDES corpora is a cornerstone of the childes-db project.
To this end we have created powerful yet easy-to-use interactive web applications that enable users to visualize various dimensions of the CHILDES corpus: frequency counts, mean lengths of utterance, type-token ratios, and more.
All of this is doable without the requirement of understanding command-line tools or any kind of programming knowledge as had been the case with CLAN.^[The LuCiD toolkit [@chang2017] provides related functionality for a number of common analyses.
In contrast to those tools, which focus on filling gaps not covered by CLAN -- e.g., the use of $n$-gram models, incremental sentence generation, and distributional word classification -- our web apps focus on covering the same common tasks as CLAN, but making the outputs into browsable visualizations.]

Our web apps are built using Shiny, an R package that enables easy app construction using `R`.
Underneath the hood, each web app is making calls to our `childesr` API and subsequently plots the data using the popular R plotting package `ggplot2`.
A user's only task is to configure exactly what should be plotted through a series of buttons, sliders, and text boxes.
The user may specify what collection, corpus, child, age range, caregiver, etc., should be included in a given analysis.
The plot is displayed and updated in real-time, and the underlying data are also available for download alongside the plot.
All of these analyses may also be reproduced using the `childesr` package, but the web apps are intended for the casual user who seeks to easily extract developmental indices quickly and without any technical overhead.
<!-- We next walk through the three web apps provided on our website and describe their functionality in more detail.  -->

@@

png::readPNG(here::here("images/freq_app.png")) %>% grid::grid.raster()
```

@@

png::readPNG(here::here("images/derived_measures_app.png")) %>% grid::grid.raster()
```

@@

png::readPNG(here::here("images/population_viewer_app.png")) %>% grid::grid.raster()
```

@@

Although the interactive analysis tools described above cover some of the most common use cases of CHILDES data, researchers interested in more detailed and flexible analyses will want to interface directly with the data in `childes-db`.
Making use of the `R` programming language [@R-base], we provide the `childesr` package. `R` is an open-source, extensible statistical computing environment that is rapidly growing in popularity across fields and is increasing in use in child language research [e.g. @song2015; @norrman2015].
The `childesr` package abstracts away the details of connecting to and querying  the database.
Users can take advantage of the tools developed in the popular `dplyr` package [@R-dplyr], which makes manipulating large datasets quick and easy.
We describe the commands that the package provides and then give several worked examples of using the package for analyses.

@@

# Using childes-db: Worked Examples

@@

One common use of CHILDES is to estimate the frequency with which children hear different words.
These frequency estimates are used both in the development of theory [e.g., frequent words are learned earlier; @goodman2008], and in the construction of age-appropriate experimental stimuli.
One benefit of the childes-db interface is that it allows for easy analysis of how the frequencies of words change over development.
Many of our theories in which children learn the structure of language from its statistical properties implicitly assume that these statistics are *stationary*, i.e. unchanging over development [e.g., @saffran1996].
However a number of recent analyses show that the frequencies with which infants encounter both linguistic and visual properties of their environment may change dramatically over development [@fausey2016], and these changing distributions may produce similarly dramatic changes in the ease or difficulty with which these regularities can be learned [@elman1993].

@@

#### Using childesr

@@

For this analysis, we will define our words of interest as the basic color words of English (except for gray, which children hear very rarely).
We store these in the `colors` variable, and then use the `get_types` function from `childesr` to get the type frequency of each of these words in all of the corpora in CHILDES.
For demonstration, we look only at the types produced by the speakers in each corpus tagged as Mother and Father.
We also restrict ourselves to children from 1--5 years old (12--60 months), and look only at the North American English corpora.

@@

color_counts <- get_types(collection = "Eng-NA", 
                          role = c("Mother", "Father"),
                          age = c(12,60), 
                          type = colors)
```

@@

To normalize correctly (i.e., to ask what proportion of the input children hear consists of these color words), we need to know how many total words these children hear from their parents in these transcripts.
To do this, we use the `get_speaker_statistics` function, which will return a total number of tokens (`num_tokens`) for each of these speakers.

@@

We now join these two pieces of information together -- how many times each speaker produced each color word, and how many total words they produced.
We then group the data into 6-month age bins, and compute the proportion of tokens that comprise each color for each child in each 6-month bin. <!-- KM: The back-to-back summarise() calls feel a little opaque. I think there is an implicit group_by() step? -->
For comparability with the web app analysis, these proportions are converted to parts per million words.

```{r join-colors-and-tokens, echo = TRUE}
count_estimates <- color_counts %>%
  left_join(parents) %>%
  mutate(age_months = target_child_age,
         age_bin = as.integer(floor(age_months / 6) * 6),
         color = tolower(gloss)) %>%
  group_by(age_bin, color, target_child_id, transcript_id) %>%
  summarise(count = sum(count), num_tokens = sum(num_tokens)) %>%
  summarise(count = sum(count), num_tokens = sum(num_tokens)) %>%  
  mutate(parts = count / num_tokens * 1e6)
```

@@

```{r color-plot, fig.height = 5, fig.width = 10, fig.cap ="Color frequency as a function of age. Points represent means across transcripts, error bars represent 95% confidence intervals computed by nonparametric bootstrap", cache = TRUE, message = FALSE, warning = FALSE}
ggplot(count_estimates_with_error, aes(x = age_bin, y = empirical_stat,
                            ymin = ci_lower, ymax = ci_upper,
                            color = color, label = color)) +
  geom_pointrange(position = position_dodge(.5)) +
  geom_point() +
  geom_smooth(method = "loess", se = F, span = .5) +
  facet_wrap(~ color, scales = "free_y") +
  theme_few() +
  scale_color_manual(values = graph_colors$graph_color) +
  scale_x_continuous(breaks = seq(12, 60, 12),
                     name = "Age (months)") +
  scale_y_continuous(name = "Frequency (per million words)") +
  geom_dl(method = list(dl.trans(x=x +.2), "last.qp", cex=1)) +
  theme(legend.position = "none")
```

@@

First, we use the `childesr` function call `get_speaker_statistics` to pull data relating to the aforementioned derived measures for children and their transcripts.
Note that we exclusively select the children's production data, and exclude their caregivers' speech.

```{r echo=TRUE}
stats <- get_speaker_statistics(role = "Target_Child")
```

@@

included_languages <- stats %>%
  filter(target_child_age < max_age * 12) %>%       
  count(language) %>% 
  filter(n > number_of_transcripts_threshold) %>%
  pull(language)
```

@@

```{r echo=TRUE}
data <- stats %>%
  filter(!is.na(target_child_sex), target_child_name != "Leo",
         language %in% included_languages) %>%
  group_by(target_child_id, target_child_age, 
           target_child_sex, language) %>%
  summarise(measure = mean(mtld)) %>%
  ungroup() %>%
  mutate(age_years = target_child_age / 12,
         target_child_sex = factor(target_child_sex, 
                                   levels = c("male","female"))) %>%
  filter(age_years < max_age)
```

@@

```{r echo=FALSE}
mod <- lmer(measure ~ target_child_sex * poly(age_years, 2) - 1 +
                    (1 | target_child_id) +
                    (target_child_sex + age_years | language),
  data = data, REML = TRUE, control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))

mod_data <- expand.grid(age_years = seq(0, max_age, .1),
                                          target_child_sex = c("male","female"),
                                          language = included_languages) %>%
  as_data_frame()

mod_data$measure <- predict(mod, newdata = mod_data,
                            re.form = ~(target_child_sex + age_years | language))
```

```{r echo=FALSE, warning=FALSE}
ggplot(data, aes(x = age_years, y = measure,
                 fill = target_child_sex, colour = target_child_sex)) +
  geom_point(alpha = .3, size=1, stroke=0) +
  geom_line(data = mod_data,
            aes(x = age_years, y = measure, color = target_child_sex)) +
  facet_wrap(~language) +
  ggthemes::theme_few() +
  ggthemes::scale_color_ptol(name = "Sex") +
  xlim(0,max_age) +
  ylim(0,15) +
  xlab("Age (years)") +
  ylab("MTLD") +
  theme(legend.position = "bottom")
```

@@

## Teaching with childes-db

@@

1. introduce the research question (e.g., What are the types of words that children first produce?),
2. allow students to reflect or do a pair-and-share discussion with their neighbor,
3. show the trajectory of a single lexical item while explaining key parts of the visualization (see Panel A of Figure \ref{fig:in-class-demo-plot}),
4. elicit hypotheses from students about the kinds of words that children are likely to produce,
5. make real-time queries to the web application to add students' suggestions and talk through the updated plots (Panels B and C of Figure \ref{fig:in-class-demo-plot}), and
6. finish by entering a pre-selected set of words that communicate the important takeaway point (Panel D of Figure \ref{fig:in-class-demo-plot}) .

```{r in-class-demo-plot, fig.pos="h", out.width = "95%", fig.cap="Worked example of using the web applications for in-class teaching. Panels A-D show how an instructor could dynamically build a plot during a lecture to demonstrate a key concept in language acquisition."}

png::readPNG(here::here("images/in_class_teach_viz.png")) %>% grid::grid.raster()
```

@@

We have presented `childes-db`, a database formatted mirror of the CHILDES dataset.
This database -- together with the `R` API and web apps -- facilitates the use of child language data.
For teachers, students, and casual explorers, the web apps allow browsing and demonstration.
For researchers interested in scripting more complex analyses, the API allows them to abstract away from the details of the CHAT format and easily create reproducible analyses of the data.
We hope that these functionalities broaden the set of users who can easily interact with CHILDES data, leading to future insights into the process of language acquisition.

childes-db addresses a number of needs that have emerged in our own research and teaching, but there are still a number of limitations that point the way to future improvements.
For example, `childes-db` currently operates only on transcript data, without links to the underlying media files; in the future, adding such links may facilitate further computational and manual analyses of phonology, prosody, social interaction, and other phenomena by providing easy access to the video and audio data.
Further, we have focused on including the most common and widely-used tiers of CHAT annotation into the database first, but our plan is eventually to include the full range of tiers.
Finally, a wide range of further interactive analyses could easily be added to the current suite of web apps.
We invite other researchers to join us in both suggesting and contributing new functionality as our system grows and adapts to researchers' needs.

# References
```{r create_r-references, warning = FALSE}
r_refs(file = "childesdb.bib")
```

>>>>

```{r load_packages, include=FALSE}
library(papaja)
library(tidyverse)
library(childesr)
library(directlabels)
library(png)
library(here)
library(lme4)
library(ggthemes)
library(optimx)
library(grid)
knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE,
                      warning = FALSE, error = FALSE)
```

@@

readPNG(here("images/schema.png")) %>% grid.raster()
```

@@

`childes-db` is stored as a MySQL database, an industry-standard, open-source relational database server that can be accessed directly from a wide range of programming languages.
The `childes-db` project provides hosted, read-only databases for direct access and for `childesr` (described below) as well as compressed .sql exports for local installation.
While the former is appropriate for most users, local installation can provide performance gains by allowing a user to access the database on their machine or on their local network, as well as allowing users to store derived information in the same database.

@@

```{r, eval=FALSE, include=FALSE}
# plot error between clan counts and our counts for speaker unigrams
freq_data <- read_csv("data/freq_testing.csv")

ggplot(freq_data %>%
         mutate(error = abs(clan - db) / clan),
       aes(error)) +
  geom_density() +
  xlim(0, .25)
```

@@

ggplot(mlu_data %>%
         mutate(error = abs(clan - db) / clan),
       aes(error)) +
  geom_density() +
  xlim(0, .25)
```

@@

# Interfaces for Accessing `childes-db`

We first discuss the `childes-db` web apps and then introduce the `childesr` R package.

@@

The ability to easily browse and explore the CHILDES corpora is a cornerstone of the `childes-db` project.
To this end we have created powerful yet easy-to-use interactive web applications that enable users to visualize various dimensions of the CHILDES corpus: frequency counts, mean lengths of utterance, type-token ratios, and more.
All of this is doable without the requirement of understanding command-line tools or any kind of programming knowledge as had been the case with CLAN.^[The LuCiD toolkit [@chang2017] provides related functionality for a number of common analyses.
In contrast to those tools, which focus on filling gaps not covered by CLAN -- e.g., the use of $n$-gram models, incremental sentence generation, and distributional word classification -- our web apps focus on covering the same common tasks as CLAN, but making the outputs into browsable visualizations.]

Our web apps are built using Shiny, an R package that enables easy app construction using R.
Underneath the hood, each web app is making calls to our `childesr` API and subsequently plots the data using the popular R plotting package `ggplot2`.
A user's only task is to configure exactly what should be plotted through a series of buttons, sliders, and text boxes.
The user may specify what collection, corpus, child, age range, caregiver, etc., should be included in a given analysis.
The plot is displayed and updated in real-time, and the underlying data are also available for download alongside the plot.
All of these analyses may also be reproduced using the `childesr` package, but the web apps are intended for the casual user who seeks to easily extract developmental indices quickly and without any technical overhead.
<!-- We next walk through the three web apps provided on our website and describe their functionality in more detail.  -->

@@

readPNG(here("images/freq_app.png")) %>% grid.raster()
```

@@

readPNG(here("images/derived_measures_app.png")) %>% grid.raster()
```

@@

readPNG(here("images/population_viewer_app.png")) %>% grid.raster()
```

@@

Although the interactive analysis tools described above cover some of the most common use cases of CHILDES data, researchers interested in more detailed and flexible analyses will want to interface directly with the data in `childes-db`.
Making use of the R programming language [@R-base], we provide the `childesr` package. R is an open-source, extensible statistical computing environment that is rapidly growing in popularity across fields and is increasing in use in child language research [e.g. @song2015; @norrman2015].
The `childesr` package abstracts away the details of connecting to and querying  the database.
Users can take advantage of the tools developed in the popular `dplyr` package [@R-dplyr], which makes manipulating large datasets quick and easy.
We describe the commands that the package provides and then give several worked examples of using the package for analyses.

@@

# Using `childes-db`: Worked Examples

@@

One common use of CHILDES is to estimate the frequency with which children hear different words.
These frequency estimates are used both in the development of theory [e.g., frequent words are learned earlier; @goodman2008], and in the construction of age-appropriate experimental stimuli.
One benefit of the `childes-db` interface is that it allows for easy analysis of how the frequencies of words change over development.
Many of our theories in which children learn the structure of language from its statistical properties implicitly assume that these statistics are *stationary*, i.e. unchanging over development [e.g., @saffran1996].
However a number of recent analyses show that the frequencies with which infants encounter both linguistic and visual properties of their environment may change dramatically over development [@fausey2016], and these changing distributions may produce similarly dramatic changes in the ease or difficulty with which these regularities can be learned [@elman1993].

@@

#### Using `childesr`

@@

For this analysis, we will define our words of interest as the basic color words of English (except for gray, which children hear very rarely).
We store these in the `colors` variable, and then use the `get_types()` function from `childesr` to get the type frequency of each of these words in all of the corpora in CHILDES.
For demonstration, we look only at the types produced by the speakers in each corpus tagged as Mother and Father.
We also restrict ourselves to children from 1--5 years old (12--60 months), and look only at the North American English corpora.

@@

color_counts <- get_types(collection = "Eng-NA",
                          role = c("Mother", "Father"),
                          age = c(12, 60),
                          type = colors)
```

@@

To normalize correctly (i.e., to ask what proportion of the input children hear consists of these color words), we need to know how many total words these children hear from their parents in these transcripts.
To do this, we use the `get_speaker_statistics()` function, which will return a total number of tokens (`num_tokens`) for each of these speakers.

@@

We now join these two pieces of information together -- how many times each speaker produced each color word, and how many total words they produced.
We then group the data into 6-month age bins, and compute the proportion of tokens that comprise each color for each child in each 6-month bin.
For comparability with the web app analysis, these proportions are converted to parts per million words.

```{r join-colors-and-tokens, echo = TRUE}
count_estimates <- color_counts %>%
  left_join(parents) %>%
  mutate(age_months = target_child_age,
         age_bin = as.integer(floor(age_months / 6) * 6),
         color = tolower(gloss)) %>%
  group_by(age_bin, color, target_child_id, transcript_id) %>%
  summarise(transcript_count = sum(count),
            transcript_num_tokens = sum(num_tokens)) %>%
  group_by(age_bin, color, target_child_id) %>%
  summarise(child_count = sum(transcript_count),
            child_num_tokens = sum(transcript_num_tokens)) %>%
  mutate(parts = child_count / child_num_tokens * 1e6)
```

@@

```{r color-plot, fig.height = 5, fig.width = 10, fig.cap ="Color frequency as a function of age. Points represent means across transcripts, error bars represent 95% confidence intervals computed by nonparametric bootstrap.", cache = TRUE}
ggplot(count_estimates_with_error, aes(x = age_bin, y = empirical_stat,
                            ymin = ci_lower, ymax = ci_upper,
                            color = color, label = color)) +
  geom_pointrange(position = position_dodge(.5)) +
  geom_point() +
  geom_smooth(method = "loess", se = F, span = .5) +
  facet_wrap(~ color, scales = "free_y") +
  theme_few() +
  scale_color_manual(values = graph_colors$graph_color) +
  scale_x_continuous(breaks = seq(12, 60, 12),
                     name = "Age (months)") +
  scale_y_continuous(name = "Frequency (per million words)") +
  geom_dl(method = list(dl.trans(x = x + .2), "last.qp", cex = 1)) +
  theme(legend.position = "none")
```

@@

First, we use the `childesr` function call `get_speaker_statistics()` to pull data relating to the aforementioned derived measures for children and their transcripts.
Note that we exclusively select the children's production data, and exclude their caregivers' speech.

```{r echo=TRUE}
speaker_stats <- get_speaker_statistics(role = "Target_Child")
```

@@

included_languages <- speaker_stats %>%
  filter(target_child_age < max_age * 12) %>%
  count(language) %>%
  filter(n > number_of_transcripts_threshold) %>%
  pull(language)
```

@@

```{r echo=TRUE}
mtld_data <- speaker_stats %>%
  filter(!is.na(target_child_sex), target_child_name != "Leo",
         language %in% included_languages) %>%
  group_by(target_child_id, target_child_age, target_child_sex, language) %>%
  summarise(measure = mean(mtld)) %>%
  ungroup() %>%
  mutate(age_years = target_child_age / 12,
         target_child_sex = factor(target_child_sex,
                                   levels = c("male", "female"))) %>%
  filter(age_years < max_age)
```

@@

```{r}
mod <- lmer(measure ~ target_child_sex * poly(age_years, 2) - 1 +
              (1 | target_child_id) +
              (target_child_sex + age_years | language),
            data = mtld_data, REML = TRUE, control = lmerControl(
              optimizer = "optimx", optCtrl = list(method = "nlminb")))

mod_data <- expand.grid(age_years = seq(0, max_age, .1),
                        target_child_sex = c("male", "female"),
                        language = included_languages) %>%
  as_data_frame()

mod_data$measure <- predict(
  mod, newdata = mod_data, re.form = ~(target_child_sex + age_years | language)
)
```

```{r, fig.asp = 0.9}
language_codes <- c("deu" = "German", "eng" = "English", "fra" = "French",
                    "jpn" = "Japanese", "nld" = "Dutch", "spa" = "Spanish",
                    "zho" = "Chinese")

plt_mtld_data <- mtld_data %>%
  mutate(language = language_codes[language])
plt_mod_data <- mod_data %>%
  mutate(language = language_codes[language])

ggplot(plt_mtld_data, aes(x = age_years, y = measure, fill = target_child_sex,
                          colour = target_child_sex)) +
  geom_point(alpha = .3, size = 1, stroke = 0) +
  geom_line(aes(x = age_years, y = measure, color = target_child_sex),
            data = plt_mod_data) +
  facet_wrap(~language) +
  theme_few() +
  scale_color_ptol(name = "Sex") +
  scale_fill_ptol(guide = FALSE) +
  xlim(0, max_age) +
  ylim(0, 15) +
  xlab("Age (years)") +
  ylab("MTLD") +
  theme(legend.position = "top",
        legend.box.spacing = unit(0, "mm"))
```

@@

## Teaching with `childes-db`

@@

1. introduce the research question (e.g., What are the types of words that children first produce?),
2. allow students to reflect or do a pair-and-share discussion with their neighbor,
3. show the trajectory of a single lexical item while explaining key parts of the visualization (see Panel a of Figure \ref{fig:in-class-demo-plot}),
4. elicit hypotheses from students about the kinds of words that children are likely to produce,
5. make real-time queries to the web application to add students' suggestions and talk through the updated plots (Panels b and c of Figure \ref{fig:in-class-demo-plot}), and
6. finish by entering a pre-selected set of words that communicate the important takeaway point (Panel d of Figure \ref{fig:in-class-demo-plot}).

```{r in-class-demo-plot, fig.pos="h", out.width = "95%", fig.cap="Worked example of using the web applications for in-class teaching. Panels a-d show how an instructor could dynamically build a plot during a lecture to demonstrate a key concept in language acquisition."}

readPNG(here("images/in_class_teach_viz.png")) %>% grid.raster()
```

@@

We have presented `childes-db`, a database formatted mirror of the CHILDES dataset.
This database -- together with the R API and web apps -- facilitates the use of child language data.
For teachers, students, and casual explorers, the web apps allow browsing and demonstration.
For researchers interested in scripting more complex analyses, the API allows them to abstract away from the details of the CHAT format and easily create reproducible analyses of the data.
We hope that these functionalities broaden the set of users who can easily interact with CHILDES data, leading to future insights into the process of language acquisition.

`childes-db` addresses a number of needs that have emerged in our own research and teaching, but there are still a number of limitations that point the way to future improvements.
For example, `childes-db` currently operates only on transcript data, without links to the underlying media files; in the future, adding such links may facilitate further computational and manual analyses of phonology, prosody, social interaction, and other phenomena by providing easy access to the video and audio data.
Further, we have focused on including the most common and widely-used tiers of CHAT annotation into the database first, but our plan is eventually to include the full range of tiers.
Finally, a wide range of further interactive analyses could easily be added to the current suite of web apps.
We invite other researchers to join us in both suggesting and contributing new functionality as our system grows and adapts to researchers' needs.

# References
```{r create_r-references}
r_refs(file = "childesdb.bib")
```
